Reformulating phylogenetic mixed models to improve flexibility and speed

### 2019-09-09 phyloglmm_ms.tex
pushing ms

\linenumbers

@@

% \pagenumbering{gobble}

@@

\begin{center}
{\Large
\textbf\newline{Reformulating phylogenetic mixed models to improve flexibility and speed} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
\vspace{2pt}
Michael Li\textsuperscript{1*} and Benjamin M. Bolker\textsuperscript{1,2,3}
\\
\vspace{2pt}
* Corresponding author: Michael Li; lim88@mcmaster.ca
\end{center}
\vspace{5pt}
\textbf{1} Department of Biology, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{2} Department of Mathematics and Statistics, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{3} Institute for Infectious Diseases Research, McMaster University, Hamilton, Ontario, Canada
\\
\bigskip
% 
% \section*{JD's Comments}
% 
% Sharpen opening framing somehow. Clarify the basic idea of
% taxonomic-based investigations. Intro could have two or three
% paragraphs, and could start by explaining either PCM or phylogenetic
% regression.
% 
% Work on topic sentences. For example, the first paragraph
% in the Challenges section has absolutely the wrong topic sentence.
% Think of a paragraph as a unit of your story: you need more paragraphs
% in many places. This can be achieved with paragraph breaks and some
% sentence-shuffling.
% 
% I frankly think you could make a lot of progress if you look at every
% section and ask: what are my paragraphs, and what should be their
% topic sentences?
% 
% Clarify the hierarchical structure of the document. It's not
% completely clear whether or not challenges is nested under
% introduction or is the same level.
% 
% In either case, some of the stuff in challenges should be postponed to
% methods. Everything pre methods should be aimed at getting across a
% high-level message of what this paper is adding.
% 
% Challenges needs to be less detailed, and more clear about the
% fundamental relationship between methods. You say your method is not
% mathematically equivalent the other methods, which sort of implies
% that they're all mathematically equivalent to each other. But this
% last fact will come as a surprise or even a confusion to the reader.
% 
% You should develop a clear point of view on the relationship between
% PCM and phylogenetic regression. Sometimes it seems like the former is
% broader than the ladder, and sometimes you seem to use them as
% synonyms. It kind of seems to me like you could  talk much less about
% PCM, I guess this is mostly a paper about phylogenetic regression. Or
% maybe it's a paper about "phylogenetic linear models", and we should
% just introduce this term?
% It kind of seems like methods needs an introductory paragraph before
% you jump into the details. I don't usually do that, but this is a
% pretty complicated method section.
% 
% You need to say how the glmm approach relates conceptually to the
% simple phylogenetic regression above it.
% 
% Be consistent about your transpose notation.
% 
% The text around (3) is quite deep. It is not clear to me what you mean
% by term-wise. You have not explained what you mean by grouping
% factors.And the J Matrix, which apparently doesn't even need to be
% square, is just an identity Matrix. There's probably a way to present
% this that would lead to better intuition.
% 
% 
% Can you say a little more about the intuition or mechanism behind
% Khatri-Rao? KR seems to work on partitioned matrices, but I don't
% see clearly how these matrices are partitioned.
% 
% The text around (5) should have less words and strive to be more
% explanatory. You can rely more on the equation.
% 
% Be consistent with the exposition of simulation models. Groups are not
% well explained, and you suddenly talk under multi-group about one
% observation per species -- presumably that was true for the previous
% model?
% 
% (6) is too clunky. Is there a way to simplify the notation for MVNs?
% Or just for the expandend covariance matrices, I guess.
% 
% Platforms is a single huge ¶ where the topic s. is actually tangential
% (could literally be removed). See if you can get clear ideas across
% and rely more on the table.
% 
% "We simulated 100 phylogenetic trees … to simulate responses …" Does
% anything after the first ellipsis carry information? What are the
% cites supporting?
% 
% "Correlation $\rho$ is used in place of covariances" -- what does this mean?
% 
% "Each realization was fitted to all model variants using a range of
% different platforms." Unnecessarily vague, tighten up here.
% 
% Be careful with the word "identical". If pez is really identical, can
% you explain briefly why and maybe say "and therefore is not shown"? Is
% that why pez doesn't appear in F2?
% 
% It should be more clear to the reader when you are using your method
% and when you are using older methods. Is F2 all implementations of the
% new method? Say so.
% 
% F3. What is meant by layout of the packages?
% 
% brms coverage is very high in F4. Are you using the right Bayesian method?
% 
% In F2 maybe use dashed boxplots for the first two and explain why.
% Also, refer to the last three in the same order in the caption as the
% legend.
% 
% I finally see from T2 that pez and phyr apparently "can't" do the
% single group. This should be briefly explained somewhere salient in
% text.
% 
% F5 and 6 captions are switched. Also, they shouldn't both refer to F2.
% Not clear that the speed one should refer to anything, since nothing
% obvious seems to be explained about the speed (nor necessarily
% needed).
% 
% F6. Why is pez fuzzed out? Does this relate to my identicality questions?
% 
% F7. This "annotations" thing is actually getting annoying. Are the
% captions just unfinished?
% 
% Yeah, I'm a stop now. You can take these comments for what they're
% worth, but I don't want to see this Chapter again until you've done
% all of the macro-editing of the Results section (figures in the right
% order, an attempt at captions, figure references correct).

@@

\begin{enumerate}
\item{Phylogenetic comparative methods (PCM) using phylogenetic regression are a powerful technique to explore relationships among related groups of species traits. However, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.}
\item{We propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but is more flexible in practice. We have implemented this formulation in two R statistical packages (\pkg{lme4} and \pkg{glmmTMB}).}
\item{Our reformulation of phylogenetic generalized linear mixed models is much more computationally efficient; it is orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
\item{Our approach can, in principle, be implemented in any platforms.}
\item{This implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to a broad range of previously difficult cases (e.g., large data, unbalanced species observations, complex random effects) and explore new ideas.}
\end{enumerate}

@@

Ever-increasing data collection capabilities (e.g., genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic databases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
Researchers use these data to fit complex models describing species occurrence and traits.
In particular, phylogenetic comparative methods (PCM) using phylogenetic regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or abundance distributions.
% \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e. how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature? \ldots}
Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate an evolutionary framework in to applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
In such cases, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships by extending exisiting software for fitting mixed effect models. 
This method allows researchers to easily incorporate evolutionary and statistical complexities without sacrificing speed.

@@

% come back to complexity
Researchers use regression-based methods such as generalized linear models (GLMs) or generalized linear mixed effect models \citep[GLMMs:][]{bolker2009generalized} to control for different levels of variation.
However, standard statistical techniques do not allow for correlations among species responses due to their shared evolutionary history.
Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}. 
If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than Brownian-motion. 
These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
Indeed, many existing methods are restricted to single observations per species, requiring users to collapse multiple observations per species to species mean values.
%% REMEMBER TO REVISIT IN DISCUSSION

@@

Classic PCMs usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects (or levels) of the predictor variables may evolve according to the phylogenetic relationship across species as well.
Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model. 
The standard PCM allow for phylogenetic correlations in the intercept of the relationship between body and brain size. 
However, species within taxonomic groups with similar body sizes may vary in overall brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have incorporated phylogenetic variation in different ways and looked at species response to phylogenetic variation with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many biologist needing to fit random-slopes model have usually turned to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
Table \ref{table:model} summarises types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.

@@

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
In particular, it easily allows our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model. 
%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2018brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
% We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
%% TODO LAST LINE SEE PAGE 5

@@

@@

\begin{equation}
\begin{aligned}
\bY & = \bX \bbeta + \besp  \\
\besp & \sim \textrm{MVN}(0,\sigma^{2} \bC), 
\label{eq:gls}
\end{aligned}
\end{equation}
where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation (PC) matrix.
The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.

@@

\subsection*{Phylogenetic generalized linear mixed model}
Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework to define a wider range that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
The generalized linear mixed effect model allows for non-Gaussian responses and uses random effects to flexibly incorporate multiple types of variability.
The typical GLMM has the form:
\newcommand{\dist}{{\cal D}}
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = \bX \bbeta + \bZ \bb  \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta))  \\
\label{eq:glmm}
\end{aligned}
\end{equation}
where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
% ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
% Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.

There are forms of random variation in the mixed model framework.
First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (which we call ``tip variation'' since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
The third type of variation that is often neglected is random slopes.
Random slopes allow fixed effects (the relationship between predictors and responses) to vary among groups.
Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).

@@

The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and a vector $\bell$ of square roots of branch lengths:

@@

\begin{equation}
\bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top}, \label{eq:ZC}
\end{equation}
where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}; $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; and $\bX_{i}$ is the $n \times p_{i}$ raw random-effects model matrix.

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7,page=1]{./figure/ssplot.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}. Total simulations $N=100$ for each category. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) fit well for all parameters. 
%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
}
\label{ssplot}
\end{figure}
\end{center}

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sstime.pdf}
  \caption{Comparison of speed for all fitting platforms: layout of the packages as in Figure \ref{ssplot}}
\label{ssplot_speed}
\end{figure}
\end{center}

>>>>

@@

\pagenumbering{gobble}
% \linenumbers

@@

\begin{center}
{\Large
\textbf\newline{Reformulating phylogenetic mixed models to improve flexibility and speed} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
\vspace{2pt}
Michael Li\textsuperscript{1*} and Benjamin M. Bolker\textsuperscript{1,2,3}
\\
\vspace{2pt}
* Corresponding author: Michael Li; lim88@mcmaster.ca
\end{center}
\vspace{5pt}
\textbf{1} Department of Biology, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{2} Department of Mathematics and Statistics, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{3} Institute for Infectious Diseases Research, McMaster University, Hamilton, Ontario, Canada
\\
\bigskip

@@

\begin{enumerate}
\item{Phylogenetic comparative methods (PCM) using phylogenetic regression are a powerful technique to explore relationships among related groups of species traits. However, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.}
\item{We propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but is more flexible in practice. We have implemented this formulation in two R statistical packages (\pkg{lme4} and \pkg{glmmTMB}).}
\item{Our reformulation of phylogenetic generalized linear mixed models is much more computationally efficient; it is orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
\item{Our approach can, in principle, be implemented on any PCM platform.}
\item{This implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to a broad range of previously difficult cases (e.g., large data, unbalanced species observations, complex random effects) and explore new ideas.}
\end{enumerate}

@@

%%% PCM accounts for evolutionary relationships. More people are doing it.
Phylogenetic comparative methods (PCM) is a powerful technique to explore relationships among related groups of species traits.
Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.

Ever-increasing data collection capabilities (e.g., genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic databases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
Researchers use these data to fit complex models describing species occurrence and traits.
For example, ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate an evolutionary framework in to applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
% \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e. how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature? \ldots}

Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
In such cases, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships by extending exisiting software for fitting mixed effect models. 
This method allows researchers to easily incorporate evolutionary and statistical complexities without sacrificing speed.

@@

% Researchers use regression-based methods such as generalized linear models (GLMs) or generalized linear mixed effect models \citep[GLMMs:][]{bolker2009generalized} to control for different levels of variation.
Standard statistical regression techniques do not allow for correlations among species responses due to their shared evolutionary history.
Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}. 
If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than Brownian-motion. 
These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
Indeed, many existing methods are restricted to single observations per species, requiring users to collapse multiple observations per species to species mean values.
%% REMEMBER TO REVISIT IN DISCUSSION

@@

Classic phylogenetic regressions usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects (or levels) of the predictor variables may evolve according to the phylogenetic relationship across species as well.
Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model. 
The standard phylognetic regression allow for phylogenetic correlations in the intercept of the relationship between body and brain size. 
However, species within taxonomic groups with similar body sizes may vary in overall brain size, or taxonomic groups may vary in the relationship between brain and body size. 
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have incorporated phylogenetic variation in different ways and looked at species response to phylogenetic variation with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many biologist needing to fit random-slopes model have usually turned to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
Table \ref{table:model} summarises types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.

@@

In this paper, we will propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
In particular, it easily allows our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model. 
%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2018brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
% We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
%% TODO LAST LINE SEE PAGE 5

@@

We generated test data using a simple framework that combines fixed effects, phylogenetic random intercept, slope, and their correlations from a phylogenetic tree, as well as residuals. 
We then fit the test data from these simulations using our approach implemented using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB} that matches the assumptions of the simulation model, as well as various simplifications with other platforms.

@@

\begin{equation}
\begin{aligned}
\bY & = \bX \bbeta + \besp  \\
\besp & \sim \textrm{MVN}(0,\sigma^{2} \bC), 
\label{eq:gls}
\end{aligned}
\end{equation}
where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic covariance (PC) matrix.
The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.

@@

\subsection*{Phylogenetic generalized linear mixed model}
Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework to define a wider range that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
The generalized linear mixed effect model allows for non-Gaussian responses and uses random effects to flexibly incorporate multiple types of variability.
The typical GLMM has the form:
\newcommand{\dist}{{\cal D}}
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = \bX \bbeta + \bZ \bb  \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta))  \\
\label{eq:glmm}
\end{aligned}
\end{equation}
where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
When $\dist$ is Gaussian, $g$ is the identity function, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
% ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
% Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.

There are forms of random variation in the mixed model framework.
First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
The third type of variation that is often neglected is random slopes.
Random slopes allow fixed effects (the relationship between predictors and responses) to vary among groups.
Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).

@@

The $\bS$ matrix is the product of an $m \times b$ indicator matrix $\bS_{ind}$ of branch indices and a vector $\bell$ of square roots of branch lengths:

@@

\begin{equation}
\bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top}, \label{eq:ZC}
\end{equation}
where $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; $\bX_{i}$ is the $n \times p_{i}$ raw random-effects model matrix; and $\ast$ is the Khatri-Rao product \citep{khatri1968solutions} partitioned at the observation level ($n$).

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7,page=1]{./figure/ssplot.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}. Total simulations $N=100$ for each category. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{lme4},  \pkg{glmmTMB}, and \pkg{brms}) fit well for all parameters. 
%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
}
\label{ssplot}
\end{figure}
\end{center}

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sstime.pdf}
  \caption{Comparison of speed for all fitting platforms.}
\label{ssplot_speed}
\end{figure}
\end{center}
